---
title: "RchivalTag Package Tutorial"
subtitle: 'supporting material for the RchivalTag package'

author:
- name: "Dr. Robert Bauer | Data Scientist & Fishery Biologist"
  affiliation: 1

address:
- code: 1
  #address: Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA 
  email: marine.biologging@gmail.com
  orcid: https://orcid.org/0000-0003-4224-8023
  
date: "February 10, 2020"
bibliography: library.bib
tags: Archival Tagging data, Satellite tags, Wildlife Computers, LOTEK


output: 
  uswebr::html_uswds:
    number_sections: FALSE
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


## Motivation and background:
The idea behind the [RchivalTag](https://github.com/rkbauer/R_Tutorial_Analyzing_Fish_Telemetry_Data_with_RchivalTag)-package and this tutorial is to facilitate the analysis and visualization of archival and pop-up archival tagging data within R.
The package (and tutorial) focuses on mainly on the following data products:

1. Read-in functions for different archival tag data products
2. Time-at-Depth (TaD) and Time-at-Temperature (TaT) Histogram data
3. Time series data (e.g. depth and temperature)
4. PAT-style Depth Temperature profiles (PDT) data
5. Tracks (Geolocation Estimates and Likelihood Areas from light levels, depth and temperature)

For more details see below as well as the package-documentation [@Bauer2020]:
```{r, collapse=TRUE}
## install or load package
# install.packages("RchivalTag")
library("RchivalTag")

## Package overview:
?RchivalTag 
help(package="RchivalTag") ## list of functions
```

## Details 

### TaD-/TaT-histogram data
- The package allows to read and generate standard summary data products (TaD-/TaT-profiles, see above) from recovered or transmitted time series data sets as well as to merge and visualize such summary data products from different tag setups/tagging programs. For more information on these data products, please see: [@WildlifeComputers2019; @Bauer2017a]

### Depth and temperature time series data
- (interactive) data visualization, optionally highlighting temperature records and daytime differences (dawn, day, dusk, night).

### Daily Depth-temperature time series data
- data visualization and examination of the thermal stratification of the water column (i.e. thermocline depth, gradient and stratification index), based on previously interpolated depth-temperature time series data or PAT Style depth temperature profiles (PDT). The paper by [@Bauer2015b] is highly recommended in this context.

### Visualizing Geolocation Estimates and Likelihood Areas
- The newest version of [RchivalTag](https://github.com/rkbauer/R_Tutorial_Analyzing_Fish_Telemetry_Data_with_RchivalTag) includes a set of functions to illustrate geolocation estimates and likelihood areas from archival tags in (interactive) figures (e.g. see ggplot_geopos(), ggplotly_geopos() and leaflet_geopos()).


## Compatability
So far, the package is mainly adapted for (pop-up) archival tagging data from [Wildlife Computers](https://wildlifecomputers.com/) and [LOTEK](https://www.lotek.com/) PSAT tags, but can also be applied to data from other tag manufacturers (e.g. see ts2histos() in order to calculate TaD & TaT-frequencies from time series data). Function examples are based on the transmitted data sets of a miniPAT-tag from the [BLUEMED-project](http://bluemed-project.com/), funded by the French National Research Agency ([ANR](http://www.agence-nationale-recherche.fr)).


## References








