---
title: "Interactive Acoustic Data Abacus Plots"

author:
  - name: "Dr. Robert Bauer | Data Scientist & Fishery Biologist"
affiliation: 1

address:
- code: 1
  email: marine.biologging@gmail.com
  orcid: https://orcid.org/0000-0003-4224-8023
  address: https://github.com/rkbauer

date: "February 10, 2020"
#bibliography: library.bib
tags: Acoustic Tagging data, radio telemetry, Vemco, LOTEK, ATS. ETN, OTN, coastal array, receiver


output: 
  #pdf_document:
  #  toc: FALSE
  uswebr::html_uswds:
    number_sections: FALSE
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

### Raw detection abacus plot
This tutorial illustrates how to create an abacus plot from raw acoustic defections from and array of several Vemco VR2W acoustic receivers.
The example data is from an ongoing project and was masked (date vectors, receiver and transmitter ids and locations have been changed) for confidentiality reasons.
It is freely available on https://github.com/rkbauer/interactive_abacus_plots_acoustic_data.
The code can be applied to several types of acoustic or radio telemetry arrays such as fish passes, rivers, estuaries, or in linear coastal arrays. 

```{r}
library(here)
load(here::here("interactive_acoustic_or_radio_telemetry_time_series_plots/raw_detections_blackwater.rd"), verbose = TRUE) 
str(df,1)
```

First we need to transform the raw detection time series into a `dygraphs` format:
```{r, collapse=TRUE}
y <- "Station.Name"
id <- "A69-9001-6765"
input <- df[which(df$Transmitter %in% id),]
input$datetime <- as.POSIXct(input$Date.and.Time..UTC.,tz="UTC")
input$date <- as.Date(input$datetime)

dat <- input[,c("datetime",y)]
dat[[y]] <- as.numeric(gsub("BW ","", dat[[y]]))

ds <- data.frame(dat[,y]); names(ds) <- y
dat_xts <- xts::xts(ds,order.by = dat$datetime)
```

Let's start plotting:

```{r}
# plot arguments:
drawPoints <- labelsUTC <- doRangeSelector <- T
pointSize <- 4; strokeWidth <- 0; col <- "blue"; label <- y

library(dygraphs)
dg <- dygraph(dat_xts)
dg <- dg %>% dyOptions(colors=col, drawPoints=drawPoints, pointSize=pointSize, 
                       labelsUTC=labelsUTC, strokeWidth = strokeWidth)
dg <- dg %>% dyAxis("y", label = "River Station")
if(doRangeSelector) dg <- dg %>% dyRangeSelector()

## the axisLabelFormatter option lets you specify a JavaScript function 
## to format the labels on axis tick marks for display.
dg <- dg %>% dyAxis(
  "y",
  axisLabelFormatter = 'function(d){return "BW" + d.toString()}',
  axisLabelWidth = 70
)

dg
```


### How to add day-night shadings
The code below is adapted from the `dy_DepthTS` function of the `RchivalTag` package which can be used to illustrate depth time series data.
In order to add day-night shadings to our figure, we need first to estimate the daily sunrise and sunset.
To do this, we calculate the average position per day via the `ddply` function and fill position gaps in the time series (no detection periods) via a regression spline.
Finally we can estimate daily sunrises and sunsets based on our positions via the `get_DayTimeLimit`s` function from `RchivalTag`

```{r}
names(input)
library(plyr)
pos0 <- ddply(unique(input[,c("date","Longitude","Latitude")]),.(date), 
              Lon=mean(x$Longitude),Lat=mean(x$Latitude))
dates <- as.Date((min(input$date)-1):(max(input$date)+1),origin="1970-01-01")
pos <- merge(pos0, data.frame(date=dates),by="date",all.y = T)
pos$Lon <- spline(1:nrow(pos), y = pos$Lon, xout = 1:nrow(pos))$y
pos$Lat <- spline(1:nrow(pos), y = pos$Lat, xout = 1:nrow(pos))$y
pos$datetime <- RchivalTag:::.date2datetime(pos$date,tz = "UTC")
head(pos)
pos <- RchivalTag::get_DayTimeLimits(pos)
```

In the next step we need transform the periods into a `dygraph` format:

```{r}
shade_periods <- c("sunrise","sunset","dawn.ast","dusk.ast")
shades <- unique(pos[,shade_periods])

shades_list <- list()
j <- 1

for(i in 1:nrow(shades)){
  add <- list(from=as.POSIXct(shades$sunrise[i],tz = "UTC"), 
              to=as.POSIXct(shades$sunset[i],tz = "UTC"),
              color="white")
  shades_list[[j]] <- add
  j <- j+1
  add <- list(from=as.POSIXct(shades$dawn.ast[i],tz = "UTC"), 
              to=as.POSIXct(shades$sunrise[i],tz = "UTC"),
              color="lightgrey")
  shades_list[[j]] <- add
  j <- j+1
  add <- list(from=as.POSIXct(shades$dusk.ast[i],tz = "UTC"), 
              to=as.POSIXct(shades$sunset[i],tz = "UTC"),
              color="lightgrey")
  shades_list[[j]] <- add
  j <- j+1
}

```

Now we can add the formated shadings to the existing `dygraph` object:

```{r}
## set background to grey
dg <- dyShading(dg, 
                from = pos$datetime[1] , 
                to = tail(pos$datetime,1),
                color = "darkgrey" )

## add twilight and daytime shadings
for( period in shades_list ) {
  dg <- dyShading(dg, 
                  from = period$from , 
                  to = period$to,
                  color = period$color)
  dg <- dyAnnotation(dg, 
                     x = mean(c(period$from,period$to)), 
                     text = period$label, 
                     attachAtBottom=T)
}

dg
```

You can further edit the object `dg` of this plot via the `dygraphs` package (https://rstudio.github.io/dygraphs/).

Enjoy coding!






